\name: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-03-04 paper/main.tex
Updated full model all layers and some TODOs in methods

\subsection{Datasets}
We obtain a dataset consisting of $\approx 11000$\todo{precise number}
human proteins with over 170.000\todo{exact number} links from STRING
\citep{STRINGv10}. For the drug-target interactions, we
fetched 137,000\todo{precise} links from the STITCH database
\citep{STITCHv5}. As both STRING and STITCH provide confidence scores
for each association, we filtered them as advised by a threshold of
$700$, therefore retaining only high-confidence interactions.

@@

We only use proteins in our analysis that have at least one link in
either STITCH or STRING, and drugs with at least one side effect and
one existing target. Therefore, the intersection between these
resources yields 1,160 drugs and 6,680 human proteins for the training
phase. We provide links to and methods for obtaining and processing
the necessary data on Github.

For comparative evaluation, we use the gold standard dataset
introduced by \citet{Yaminishi2008} \todo[inline]{consisting of XXXX
  interactions between XXX drugs and XXX protein targets}.

@@

\subsubsection{Siamese neural networks and modular learnable feature transformation}

\todo[inline]{Looking at the code, I wonder if what is implemented is
  actually a Siamese ANN; it does not look as if the weights are
  shared between the two networks, so looks more like a twin or
  half-twin network? Siamese would be sim(f(e1),f(e2)) but the
  implementation seems to be sim(f(e1), g(e2)).}

As we want to learn from the similarity of drug side effects and
protein phenotypes, we use a deep Siamese neural network with
a contrastive loss using cosine similarity.
A Siamese neural network aims to learn a similarity between two
embeddings.

@@

Therefore, the precomputed embeddings are run through a regular,
neural learnable feature transformation (LFT) network, which also
reduces the eventual representation size for drugs and proteins
separately. An example structure for both types of features can be
found in Figure~\ref{fig:SiameseNetwork}.

@@

\begin{figure}[!tpb]%figure1
	\centerline{\includegraphics[width=0.35\textwidth]{figures/siamese_network.png}}
	\caption{Siamese network applied to molecular and DL2vec
          features, utilizing deep learnable feature transformations
          (LFT). The similarity function $\otimes$ yields the
          similarity between both transformed embeddings e.g. by
          computing the cosine similarity.}
	\label{fig:SiameseNetwork}
\end{figure}

@@

\subsubsection{Hyperparameter tuning}
As the number of drug-targets are sparse with respect to the number of
both drugs and proteins considered, the training, validation and
testing datasets are imbalanced. As there are only $22,336$ links in
the considered subset the ratio
\begin{equation*}
	\frac{\#drugs \cdot \#proteins}{\#dti\_links} \approx 360,
\end{equation*}
consequently needing compensation in the computed loss function and
appropriate metrics for the evaluation.

Therefore, we weighted all positive drug-protein pair samples with
this ratio by introducing the following loss function with respect to
binary cross-entropy:
\begin{equation}
	l(x,y) = - w \left[ y \cdot \log x + (1 - y) \cdot \log (1 - x) \right]
\end{equation}
for a given prediction $x$ and target $y$. We average this loss among
all drug-protein pairs in the training set, leading to a stable
environment for the used optimizing scheme \textit{Adam}
\citep{Adam2014}. We implemented a 5-fold cross validation among the
proteins. Furthermore, we used early stopping to detect plateaus in
the training process.

To find the best hyperparameter configuration for the proposed model
we performed a grid search to find the most expressive and
non-redundant representation. We pretrained the bottom-up and the
top-down model separately and aimed at best performing models
with respect to our evaluation metrics. We optimized embedding sizes,
depth of the neural network, optimizer, learning rate and layer types
using an extensive, manual grid search. \todo[inline]{Specify range of
  values searched here; even better if you have the intermediate
  results and can put them here (better: in the supplement).}

@@

\todo[inline]{Needs a statement on how the PAIRS are ranked here and
  TPR/FPR is calculated based on the ranks of the DT-PAIRS.}
We calculate the AUROC by computing true positive rate at various
false positive rate thresholds and using trapezoidal approximations to
estimate the area under the curve. We refer to this measure as
$MacroAUC$.

\todo[inline]{For RH: check this again later:}
In contrast we also calculate the $MicroAUC$ score. For given lists $D, P$ of drugs and proteins, respectively, and a set of known interactions $Int := \{(d_i, p_i) \}$, \textit{MicroAUC} is calculated as the average per entity (macro) \textit{AUROC} score. With respect to proteins, this can be formalized for given labels and prediction $l,y:D\times P \rightarrow \{0,1\}$  as 
\begin{equation*}
	MicroAUC_p := \underset{p\in P}{mean}\left(\left\{ \text{AUROC}(\{ (l(d_i, p), y(d_i,p))| d_i\in D\}) \right\}\right)
\end{equation*}

Drugs and proteins can be interchanged in this formulation, and we
refer to the different measures as a protein-centric microAUC
($MicroAUC_p$) and a drug-centric microAUC
($MicroAUC_d$). Additionally, the $MicroAUC$ score may not be defined
as, in some datasets some targets or drugs, respectively, do not have
any interactions, leading to an infeasible $TPR$ and an undefined
$AUROC$ score for that entity.
\todo[inline]{Don't understand this, can you add the formula:} For
those entities we impute the $MicroAUC$ linearly, by using the
accuracy for this subset.

@@

\todo[inline]{Introduction}
Within DTI prediction, there are potential biases resulting from the
underlying datasets \citep{Pahikkala2014}. First, novel drugs are
often designed by altering non-functional components of a drug,
leading to two and more very similar drugs designed to target the same
proteins \cite{Overington2006}. This can result in a bias when it
leads to hidden duplicates that can distribute among the train/test
split, resulting in a better (measured) predictive performance than
would be expected when the model is applied to identify drugs that
target a protein for which no drugs yet exist. Second, some proteins
(which we call \textit{hub proteins}) have significantly more known
interactions with drugs than others. In the STITCH database, $5\%$ of
the proteins have $40\%$ of the interactions, and similar
distributions are present in the Yamanishi and Drugbank
\cite{Drugbank2007, Drugbank2017} datasets; preferentially predicting
these proteins may increase predictive performance while not
reflecting the actual performance when applied to a new protein (e.g.,
a protein for which no interactions are known). These differences in
the number of drugs targeting certain proteins may be the result of
study bias where more ``valuable'' proteins have more drugs designed
to target them due to their involvement in more common diseases (or
diseases for which drugs can be more profitably marketed).

We developed \name as a computational model to predict drug--target
interactions. Specifically, given a protein, \name will identify and
rank drugs that likely target this protein. \name combines two types
of features: structural information for drugs and proteins that can be
used to determine if the drug and protein physically interact, and
information about phenotypic effects of drugs and changes in protein
function that may ``localize'' on an interaction network (i.e.,
neighboring nodes will share some of these features or be
phenotypically similar).  As structural features, \name uses
structural representations of drugs from the SMILES transformer
\cite{SmilesTransformer} and representations of protein amino acid
sequences from DeepGOPlus \cite{DeepGoPlus}.  \name learns
representations of drug effects and protein functions using the
ontology-based machine learning method DL2Vec \cite{DL2vec2020} and
ontology-based annotations of drugs and proteins.

@@

We use two Siamese neural networks to combine the molecular and
phenotype representations of drugs and proteins. Siamese neural
networks are used to learn a kind of similarity between the molecular
and phenotype representations. We add these similarities to a protein
interaction network as node labels; for each drug, proteins receive
two features (from the molecular and phenotype similarities of the
drug to the protein).
\todo[inline]{There is a missing sentence here explaining the graph
  neural network, output, etc.}
We then train this model in an end-to-end manner using a graph neural
network architecture.  The overall workflow\todo{make sure it's a
  workflow} is depicted in Figure~\ref{fig:Fullmodel}.

@@

\begin{figure}[!tpb]
	\centerline{\includegraphics[width=0.45\textwidth]{figures/full_model.png}}
	\caption{Full DTI prediction model based on the pretrained
          learnable feature transformations (LFT) for either molecular
          structure or ontology based features. The transformed
          protein representations are added to each corresponded
          protein as node features for the graph convolutional
          steps.}
	\label{fig:Fullmodel}
\end{figure}

@@

We evaluate our model's ability to identify drug--target interactions
using different approaches and datasets.  First, we perform a
cross-validation over proteins and validate our results. A
cross-validation over proteins aims to evaluate how the model performs
when tasked to identify drugs that may target a ``novel'' protein,
i.e., one not seen during training, or a protein for which a drug that
targets it should be predicted.  \todo[inline]{Describe/add the CV
  results here.}

Our model is trained on the STITCH database. We use the model trained
on STITCH and evaluate the model on the Yamanishi benchmark
dataset\todo{This needs another sentence stating how you made sure
  there was no overlap in data.}.

>>>>

\subsection{Datasets}
We obtain a dataset consisting of 12884 human proteins with over 340627 links from STRING
\citep{STRINGv10}. For the drug-target interactions, we
fetched 229870 links from the STITCH database
\citep{STITCHv5}. As both STRING and STITCH provide confidence scores
for each association, we filtered them as advised by a threshold of
$700$, therefore retaining only high-confidence interactions.

@@

We only use proteins in our analysis that have at least one link in
either STITCH or STRING, and drugs with at least one side effect and
one existing target. Therefore, the intersection between these
resources yields 1,428 drugs and 7,368 human proteins with 32212 interaction links for the training phase. We provide links to and methods for obtaining and processing
the necessary data on Github.

For comparative evaluation, we use the gold standard dataset
introduced by \citet{Yaminishi2008} consisting of 1923 interactions between 708 drugs and 1512 proteins.

@@

\subsubsection{Half-twin neural networks and modular learnable feature transformation}

\todo[inline]{Looking at the code, I wonder if what is implemented is
  actually a 
   ANN; it does not look as if the weights are
  shared between the two networks, so looks more like a twin or
  half-twin network? Siamese would be sim(f(e1),f(e2)) but the
  implementation seems to be sim(f(e1), g(e2)).}

As we want to learn from the similarity of drug side effects and
protein phenotypes, we use a deep half-twin neural network with
a contrastive loss using cosine similarity.
A half-twin neural network aims to learn a similarity between two
embeddings.

@@

Therefore, the precomputed embeddings are run through a regular,
neural learnable feature transformation (LFT) network, which also
reduces the eventual representation size for drugs and proteins
separately. An example structure for both types of features can be
found in Figure~\ref{fig:HalfTwinNetwork}.

@@

\begin{figure}[!tpb]%figure1
	\centerline{\includegraphics[width=0.35\textwidth]{figures/siamese_network.png}}
	\caption{Half-twin network applied to molecular and DL2vec
          features, utilizing deep learnable feature transformations
          (LFT). The similarity function $\otimes$ yields the
          similarity between both transformed embeddings e.g. by
          computing the cosine similarity.}
	\label{fig:HalfTwinNetwork}
\end{figure}

@@

\begin{figure}[!tpb]%figure1
	\centerline{\includegraphics[width=1\columnwidth]{figures/full_model_all_layers.png}}
	\caption{Residual architecture built by \citet{DeepGCN2019} and \citet{DeeperGCN2020} enabling deeper graph convolutional models}
	\label{fig:FullModelAllLayers}
\end{figure}

\subsubsection{Hyperparameter tuning}
As the number of drug-targets are sparse with respect to the number of
both drugs and proteins considered, the training, validation and
testing datasets are imbalanced. As there are only $22,336$ links in
the considered subset the ratio
\begin{equation}
	w:= \frac{\#drugs \cdot \#proteins}{\#dti\_links} \approx 360,
	\label{weight}
\end{equation}
consequently needing compensation in the computed loss function and
appropriate metrics for the evaluation.

Therefore, we weighted all positive drug-protein pair samples with
this ratio by introducing the following loss function with respect to
binary cross-entropy:
\begin{equation}
	l(x,y) = - w \left[ y \cdot \log x + (1 - y) \cdot \log (1 - x) \right]
\end{equation}
for a given prediction $x$ and target $y$, and positive weight $w$ defined by equation \eqref{weight}. We average this loss among
all drug-protein pairs in the training set, leading to a stable
environment for the used optimizing scheme \textit{Adam}
\citep{Adam2014}. We implemented a 5-fold cross validation split among the
proteins. Furthermore, we used early stopping in
the training process for optima detection.

To find the best hyperparameter configuration for the proposed model
we performed a grid search to find the most expressive and
non-redundant representation. We pretrained the bottom-up and the
top-down model separately and aimed at best performing models
with respect to our evaluation metrics. We optimized embedding sizes,
depth of the neural network, optimizer, learning rate and layer types
using an extensive, manual grid search, Starting from na\"ive, shallow feature transformations with an embedding size of 10, we scaled the network up to residual structures with up to 10 hidden layers leading to embeddings of size 4000, testing different network widths and learning rates for each configuration.

@@

We calculate the AUROC by computing true positive rate at various
false positive rate thresholds and using trapezoidal approximations to
estimate the area under the curve. We refer to this measure as
$MacroAUC$.

\todo[inline]{For RH: check this again later:}
In contrast we also calculate the $MicroAUC$ score. For given lists $D, P$ of drugs and proteins, respectively, and a set of known interactions $Int := \{(d_i, p_i) \}$, \textit{MicroAUC} is calculated as the average per entity (macro) \textit{AUROC} score. With respect to proteins, this can be formalized for given labels $l:D\times P \rightarrow \{0,1\}$ and predictions $y:D\times P \rightarrow [0,1]$  as 
\begin{equation*}
	MicroAUC_p := \underset{p\in P}{mean}\left(\left\{ \text{AUROC}(\{ (l(d_i, p), y(d_i,p))| d_i\in D\}) \right\}\right)
\end{equation*}

Drugs and proteins can be interchanged in this formulation, while we
refer to the different measures as protein-centric microAUC
($MicroAUC_p$) and a drug-centric microAUC
($MicroAUC_d$). Additionally, the $MicroAUC$ score may not be defined
as in some datasets some targets or drugs, respectively, do not have
any interactions, leading to an infeasible $TPR=0$ for all thresholds and an undefined
$AUROC$ score for that entity. For those entities we impute the $MicroAUC$ interpolating linearly, by using the accuracy for this subset.

@@

Within DTI prediction, there are potential biases resulting from the
underlying datasets \citep{Pahikkala2014}. First, novel drugs are
often designed by altering non-functional components of a drug,
leading to two and more very similar drugs designed to target the same
proteins \cite{Overington2006}. This can result in a bias when it
leads to hidden duplicates that can distribute among the train/test
split, resulting in a better (measured) predictive performance than
would be expected when the model is applied to identify drugs that
target a protein for which no drugs yet exist. Second, some proteins
(which we will denote as \textit{hub proteins}) have significantly
more known interactions with drugs than others. In the STITCH
database, $5\%$ of the proteins have $40\%$ of the interactions, and
similar distributions are present in the Yamanishi and Drugbank
datasets \cite{Drugbank2007, Drugbank2017} datasets; preferentially
predicting these proteins may increase predictive performance while
again not reflecting the actual performance when applied to a new
protein (e.g., a protein for which no interactions are known). These
differences in the number of drugs targeting certain proteins may be
the result of study bias where more ``valuable'' proteins have more
drugs designed to target them due to their involvement in more common
diseases (or diseases for which drugs can be more profitably
marketed).

We developed \name as a computational model to predict drug--target
interactions. Specifically, given a protein, \name will identify and
rank known drugs that likely target this protein. \name combines two
types of features: structural information for drugs and proteins that
can be used to determine if they may physically interact, and
information about drug effects that may localize on an interaction
network.  As structural features, \name uses structural
representations of drugs from the SMILES transformer
\cite{SmilesTransformer} and representations of protein amino acid
sequences from DeepGOPlus \cite{DeepGoPlus}.  \name learns
representations of drug effects and protein functions using the
ontology-based machine learning method DL2Vec \cite{DL2vec2020}.

@@

We use two half-twin neural networks to combine the molecular and
phenotype representations of drugs and proteins, and add these to a
protein interaction network as node labels.  We then train this model
in an end-to-end manner using a graph neural network architecture.
The overall workflow\todo{make sure it's a workflow} is depicted in
Figure~\ref{fig:Fullmodel}.

@@

\begin{figure}[!tpb]
	\centerline{\includegraphics[width=0.45\textwidth]{figures/full_model.png}}
	\caption{Full DTI prediction model based on the pretrained learnable feature transformations (LFT) for either molecular structure or ontology based features. The transformed protein representations are added to each corresponded protein as node features for the graph convolutional steps. }
	\label{fig:Fullmodel}
\end{figure}

@@

We first perform a cross-validation over proteins and validate our
results. Cross-validation over proteins aims to evaluate how the model
performs when tasked to identify drugs that may target a ``novel''
protein, i.e., one not seen during training.
\todo[inline]{Describe/add the CV results here.}

Second, we use the model trained on STITCH and evaluate the model on
the Yamanishi benchmark dataset\todo{This needs another sentence
  stating how you made sure there was no overlap in data.}.
